---
title: "Session 4"
author: "Tim Riffe"
date: "11/12/2020"
output:   
  html_document:
    number_sections: yes
    toc: yes
params:
  output_dir: "../EDSD2020data/docs"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=6cm, ]{assets/MPIDR_square_color.pdf}\\[\bigskipamount]}
- \posttitle{\end{center}}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective

This tutorial will demonstrate how to toggle your data between long and wide formats. In a few different ways. This task is called *pivoting*. 

# Pivot concept

Here's a table example just to make the concept clear. It will also now serve as the first example. Examine the following 3 matrices:

```{r,message=FALSE, warning=FALSE, include = FALSE}
library(tidyverse)
Long <- data.frame(Var1 = c(rep("a",5),rep("b",5)),
                   Var2 = c(1:5,1:5),
                   value = runif(10))
Wide1 <- Long %>% pivot_wider(names_from=Var1,values_from=value)
Wide2 <- Long %>% pivot_wider(names_from=Var2,values_from=value)
```

A longer representation of the toy is this: It's also *tidy* in the sense we've been working with.
```{r}
Long
```

There are in this case two plausible and one absurd *wide* representations of the data. They are redundant in terms of their information content. Observe:

Spread out `Var1`
```{r}
Wide1
```

Spread out `Var2`

```{r}
Wide2
```

Often data is delivered by statistical offices in the form of spreadsheets that follow some version of wide formatting. But we've made some decent arguments in this module for using *tidy* formatting for your data processing workflows. Here's how to to move from wide to long:

The function to use is `pivot_longer()`, where the first argument (after the incoming data) is the column range, `names_to` tells it what to call the column where we stash the former column names. `values_to` tells it what to call the column where the collected values will be stored.
```{r}
Wide1 %>% 
  pivot_longer(a:b, 
               names_to = "Var1", 
               values_to = "value")
```

If you want the rows sorts as in the original `Long` object, use `arrange(Var1,Var2)` at the end of the pipeline. If you also want the columns in that same order, use `select(Var1, Var2, value)` after that. Neither of these things are necessary for most hypothetical downstream calculations, however. To shift the other wide representation to long is essentially the same, except you'd assign to a different column name. 

To *pivot* from long to wide works like so:

```{r}
Long %>% 
  pivot_wider(names_from = "Var1", values_from = "value")
```

It'd work the same for `Var2` as well. Now let's do a contrived example using real data.

## Denmark weekly COVID data

First we'll use our merged output from the Denmark merge solution given from the previous Wednesday session (session 3). This is a contrived case: say we

We first read in the *tidy* output from the previous merge exercise. Since the `csv` output was committed to the github repo, we're able to read it directly into the `R` session, with no manual download. Just click on the file in `github`, and then click the `raw` button. That url is then copied and pasted into `R` like here:
```{r}
# COV <- readRDS("Data/DNK_STMF_COV.rds")
COV <- read_csv("https://raw.githubusercontent.com/timriffe/EDSD2020data/main/Data/DNK_STMF_COV.csv")

```

Scenario: Say you want to spread weeks out over the columns in a spreadsheet to share with someone that says they need it that way. In this case, we probably want to take our different measures `cov_deaths_wk`, `cov_cases_wk`, `cov_tests_wk`, `Deaths` and stack them. We start with them in columns, so if we were to already spread out weeks over columns, we'd need to do it for each variable. That's very wide, very impractical. Instead we want weeks over columns, but the different measures stacked. 

The first step is to take care of the measure stacking. This is a `pivot_longer()` operation. We then throw out `NA` values, and are ready to `pivot_wider()` to spread the weeks over columns. Finally, we sort, but this is just aesthetic. Always with `arrange()` start with the *outermost* sort and work your way *in*. `arrange(Sex, variable, Age)` says sort `Age` within `variable` (now stacked) within `Sex`.

```{r}
WeeksInColumns <-
  COV %>% 
  pivot_longer(cov_deaths_wk:Deaths, # column range
               names_to = "variable",
               values_to = "value") %>% 
  filter(!is.na(value)) %>% 
  pivot_wider(names_from = Week,
              values_from = value) %>% 
  arrange(Sex, variable, Age) 

# just showing the top of the first 10 columns
head(WeeksInColumns)[,1:10] 
```

If instead you wanted to put `Sex` in columns it would be:

```{r}
SexInColumns <-
  COV %>% 
  pivot_longer(cov_deaths_wk:Deaths, # column range
               names_to = "variable",
               values_to = "value") %>% 
  filter(!is.na(value)) %>% 
  pivot_wider(names_from = Sex,
              values_from = value) %>% 
  arrange(variable, Week, Age)

head(SexInColumns)
```

## WPP2019 example

The `wpp2019` package (we used it earlier for the Philippines), comes in a wide format, with years spread over columns. Often, this is how spreadsheets are downloaded from Eurostat, the WHO, etc, so it's a quite common situation to encounter. Here only a single pass with `pivot_longer()` is needed to get to tidy format. Note we put the column names used for the column range in back ticks. If you don't use back ticks it will try to interpret those years as index positions, which in this case don't exist. 

```{r}
library(wpp2019)
data(popM)
popMtidy <-
  popM %>% 
  mutate(Sex = "m") %>% 
  pivot_longer(`1950`:`2020` ,
              names_to = "Year", 
              values_to = "Population") 
head(popMtidy)
```

Note that the data came out sorted not necessarily how we would have done it. `arrange()` will not to the right thing with `age` here: it will alphabetize the ages rather than put them in ascending order. Observe:

```{r}
popMtidy %>% 
  arrange(name, Year, age)
```
To get them sorted properly, the best is to recode `age` to the integer lower bounds of each age group, then sorting is a breeze. See the example from Wednesday for this.

## Further example of pivoting

You can see pivoting used in the fertility example from the day 1 
[UN fertility example](Day1#examples)

We did not work through this in class, but it's worth working through the pieces of that pipeline to understand what's happening to the data at each step.












